---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# quadtree

This package provides functionality for working with raster-like quadtrees.

## Installation

The package can be installed with the following R command:

``` r
devtools::install_gitlab("dafriend/quadtree")
```

For documentation of all the functions, see the PDF file stored at the root directory.

## Example

A quadtree object is created from a raster or matrix:

```{r example_create, fig.width=10, fig.height=5, out.width="100%"}
library(quadtree)
library(sp)
library(raster)

data(habitat, package="quadtree") #load sample data
qt = qt_create(habitat, split_threshold=.03, split_method="sd") #create a quadtree

par(mfrow=c(1,2), mar=c(3,2,2,2))
plot(habitat, zlim=c(0,1), main="raster representation", axes=FALSE, box=FALSE)
qt_plot(qt, crop=TRUE, na_col=NULL, border_lwd=.2,xlab="", ylab="", legend=FALSE, zlim=c(0,1), main="quadtree representation", axes=FALSE)
```

The package allows for flexibility in the quadtree creation process. Several functions defining how to split and aggregate cells are provided, and custom functions can be written for both of these processes. In addition, quadtrees can be created using other quadtrees as "templates", so that the new quadtree has the identical structure as the template quadtree.

Once created, cell values can be extracted, as with a raster:

```{r example_extract}
pts = cbind(c(20000,32000,5000),c(10000,30000,27000))
qt_extract(qt,pts)
```

In addition, functions are provided for calculating least-cost paths using the quadtree as a resistance surface.

```{r example_lcp, fig.width=7, fig.height=7, out.width="70%", out.height="70%"}
start_point = c(6989,34007)
end_point = c(33015,38162)
lcp_finder = qt_lcp_finder(qt, start_point)
lcp = qt_find_lcp(lcp_finder, end_point, use_original_end_points = TRUE)

qt_plot(qt, border_col="gray70", crop=TRUE, na_col=NULL, border_lwd=.5)
lines(lcp[,1:2])
points(rbind(start_point, end_point), col="red", pch=16)
```

## Learning how to use the `quadtree` package

Currently there are no vignettes, as this package is still in development, although in the future I hope to add some. However, all of the functions in the package are well documented and have extensive examples that the user can run. To understand the basics of the library, I'd recommend first reading the documentation for `qt_create` (i.e. run `?qt_create` in the console) first - this help file contains all the basic info on what a quadtree is and how it is constructed. To learn more about the least-cost path functionality, start with the documentation for `?qt_lcp_finder` and also take a look at `?qt_find_lcp` and `?qt_find_lcps`.

Most of the other functions are quite simple - take a look at the PDF manual included in the repository for information on how to use them. 

## File structure

The code here *mostly* conforms to the standard R package structure. The exception is the 'other_files' folder.

* /R - contains the R files
* /data - contains Rdata files (.rda) containing sample data
* /man - contains the documentation files
* /other_files - contains 'scratchwork' scripts that I use to test the package functionality. This folder is ignored when the package is built.
* /src - contains the C++ code
* /tests - contains unit tests

## Implementation Details

The bulk of the code is written in C++ and interfaced with R via Rcpp.

The overall design philosophy was to keep the core C++ code completely independent from the R code (i.e. no Rcpp-related code in the core C++ files.) This results in a three-tiered organization of the code - core C++ code, Rcpp C++ code, and  R code.

### Core C++ code

This consists of the following files (only the .h files are listed to avoid redundancy, but each of these files has a corresponding .cpp file):

* Matrix.h - Defines the Matrix class implementing basic matrix funcitonality
* Node.h - Defines the Node class, which are the 'nodes' of the quadtree
* Point.h - Defines a simple Point class
* PointUtilities.h - Defines a namespace containing functions for performing  calculations with Point objects
* Quadtree.h - Defines the Quadtree class, which can be seen as a wrapper that provides a link to the interconnected nodes that make up the quadtree
* ShortestPathFinder.h - Defines a class for finding the least-cost paths using a quadtree as a cost surface

As mentioned before, these files are completely independent of R and can be built and run independently of R.

### Rcpp C++ code

These files are called 'wrappers' - essentially they each contain an instance of relevant object, and provide additional Rcpp-related functions that can be accessed from R. These essentially provide the "bridge" that allows the functionality in the core C++ files to be accessed from R.

* NodeWrapper.h - wrapper class for 'Node'
* QuadtreeWrapper.h - wrapper class for 'Quadtree'
* ShortestPathFinderWrapper.h - wrapper class for 'ShortestPathFinder'
* R_interface.h - defines a namespace that currently contains only a single function, which converts an Rcpp matrix to the Matrix class I created. This function is separate from the other files because it is a general-purpose function and thus didn't fit in any of the wrapper classes.

### R code

This category consists of the files in the 'R' folder. Many of these functions are actually unnecessary, as all of the relevant functionality can be accessed from R via the 'Wrapper' classes. However, because of the object-based representation of these objects, the syntax for accessing this functionality differs from typical R syntax. Thus, most of the R functions provided here are syntactic sugar designed to make the functionality available in a format that conforms to traditional R code syntax. The two exceptions are `qt_create.R` and `qt_plot.R` which are more complex and are not simple wrappers for C++ functions.

## Notes on package structure

The 3-tiered structure described here may be needlessly complicated. I designed it this way because I am using the functionality implemented in the C++ files for a pure-C++ implementation of an agent-based model. Thus, to avoid having to unnecessarily include R interface code in those files, I wanted a way to be able to use the original files for both projects (this package and the C++ ABM) without needing to have slightly different implementations to fit their respective purposes. Thus, I settled on using the 'wrapper class' approach as a way to augment the original classes with the necessary Rcpp functionality without having to make any changes to the underlying C++ code. 

I added most of the R code later because I wanted a way to access the functions in a way that more closely resembles typical R syntax - this will hopefully make it more intuitive for R users to work with.

Overall, the structure works fairly well, but I wonder if it's overly complicated. I'm worried my desire to keep the C++ code independent of the R code results in a cumbersome project organization. Regardless, it works, and has allowed me to easily make edits to the underlying C++ structure and then use those exact files for both projects I'm working on rather than needing slightly different versions for each. There definitely might be more sophisticated ways of achieving the same purpose, but what I'm doing works.